---
title: "SIR simulation"
author: "Spencer Wadsworth"
date: "2023-02-25"
output: html_document
---

The book Stochastic Modeling for Systems Biology (SMSB) by Darren J. Wilkinson
discusses the models of and how 
to simulate from chemical and biological kinetic systems through the use of 
stochastic 
differential equation (SDE) modeling. Chapter 6 details how this is done using
the Gillespie algorithm and chapter 7 provides a handful of relevant chemical 
and biological examples. 

Though not strictly a chemical kinetics problem, the 
SIR problem is modeled very similary and the principles in SMSB may apply 
directly to sampling from and/or modeling the SIR process, see also 
\textit{The SIR dynamic model of infectious disease transmission and its 
analogy with chemical kinetics}(Simon 2020).

### Setting up the SIR compartment model

In the SIR model, there are three compartments: S representing the proportion of
a population susceptible to a disease, I the proportion currently infected, and
R the proportion recovered from the disease. The reaction equations for an SIR
system are as follows

$$ S + I \longrightarrow 2I \\ $$

$$I \longrightarrow R$$

For continuity from SMSB and Simon, I will use square brackets $[\cdot]$ to 
represent the concentration of a compartment. For the SIR model, we have the
ordinary differential equations (ODE)s 

$$\frac{d[S]}{dt} = -\beta[S][I] \\$$
$$\frac{d[I]}{dt} = \beta[S][I] -\gamma[I] \\$$
$$\frac{d[R]}{dt} = \gamma[I]$$
where $\beta$ is essentially the rate of an infected person infecting a 
susceptible person for a given time unit and $\gamma$ is the amount of time a 
person is infected before recovery or death.


$$
\frac{d}{dt}
\begin{pmatrix}
[S] \\ [I] \\ [R]
\end{pmatrix}
=
\begin{pmatrix}
-1 \;\;0 \\ 1  -1 \\
0 \;\; 1
\end{pmatrix}
\begin{pmatrix}
\beta[S][I] \\ \gamma[I]
\end{pmatrix}
$$
The matrix of $1s$ and $0s$ is the stoichiometry matrix as defined in SMBS and 
which I will derive here by use of a Petri net (def 2.3 SMSB). For the Petri
net $(P, T, Pre, Post, M)$

$$P = 
\begin{pmatrix}
S \\ I \\ R
\end{pmatrix}, \;\;\;\;
T =
\begin{pmatrix}
\text{Infection} \\ \text{Recovery}
\end{pmatrix},
$$
$$
Pre = 
\begin{pmatrix}
1 \;\; 1 \;\; 0 \\ 0 \;\; 1 \;\;0
\end{pmatrix}, \;\;
Post = 
\begin{pmatrix}
0 \;\; 2 \;\; 0 \\ 0 \;\; 0 \;\;1
\end{pmatrix}, \;\;
M = 
\begin{pmatrix}
S0 \\ I0 \\ R0
\end{pmatrix}
$$

so the stoichiometry matrix 

$$
(Post - Pre)' =
\begin{pmatrix}
-1 \;\;0 \\ 1  -1 \\
0 \;\; 1
\end{pmatrix}
$$
as seen above.

```{r}
Pre <- rbind(c(1,1,0),c(0,1,0))
Post <- rbind(c(0,2,0),c(0,0,1))

t(Post - Pre)
```

```{r}
sir <- function(x, beta = 1/2, gamma = 1/4) {
  c(-beta*x[1]*x[2],
    beta*x[1]*x[2] - gamma*x[2],
    gamma*x[2])
}

euler <- function(t=50,dt=0.001,fun=f,ic=c(1,1),...)
{
  p=length(ic)
  n <- t/dt
  xmat=matrix(0,ncol=p,nrow=n)
  x=ic
  xmat[1,]=x
  for (i in 2:n) {
    x = x + fun(x,...)*dt
    xmat[i,]=x
  }
  ts(xmat,start=0,deltat=dt)
}

plot(euler(t=40,fun=sir,ic=c(.9,.1,0)))
```

### Mass-action stochastic kinetics of SIR model

Here we have the three compartments $S, I, R$ and the two reactions we will call
$R_1$ and $R_2$, these two sets representing $P$ and $T$ defined earlier. The
two rate reactions will be 

$$
h_1(x,c_1) = c_1x_1x_2 \\
h_2(x,c_2) = c_2x_2
$$
where $c_1$ is the hazard of an infection reaction happing and $c_2$ is the 
hazard of a recovery/death. 

```{r}
gillespie <- function(N,n,...) {
  tt=0
  x=N$M
  S=t(N$Post-N$Pre)
  u=nrow(S)
  v=ncol(S)
  tvec=vector("numeric",n)
  xmat=matrix(0,ncol=u,nrow=n+1)
  xmat[1,]=x
  for (i in 1:n) {
    h=N$h(x,...)
    tt=tt+rexp(1,sum(h))
    j=sample(v,1,prob=h)
    x=x+S[,j]
    tvec[i]=tt
    xmat[i+1,]=x
  }
  return(list(t=tvec,x=xmat))
}
```

```{r}
N = list()
N$M = c(9000,10,1)
N$Pre = Pre
N$Post = Post
N$h = function(y, th = c(.008,2)) 
{return(c(th[1]*y[1]*y[2], th[2]*y[2]))}

out=gillespie(N,20000)
op=par(mfrow=c(2,2))
plot(stepfun(out$t,out$x[,1]),pch="")
plot(stepfun(out$t,out$x[,2]),pch="")
plot(stepfun(out$t,out$x[,3]),pch="")
# plot(out$x,type='l')
# par(op)

```

```{r}
discretise <- function(out, dt=1, start=0) {
  events = length(out$t)
  end = out$t[events]
  len = (end - start)%/%dt+1
  x = matrix(0,nrow=len,ncol=ncol(out$x))
  target = 0
  j = 1
  for (i in 1:events) {
    while (out$t[i] >= target) {
      x[j,] = out$x[i,]
      j = j + 1
      target = target + dt
    }
  }
  ts(x, start = 0, deltat = 0)
}

plot(discretise(out,dt=0.01))

N = list()
N$M = c(50,100)
N$Pre = matrix(c(1,0,1,1,0,1),ncol=2,byrow=TRUE)
N$Post = matrix(c(2,0,0,2,0,0),ncol=2,byrow=TRUE)
N$h=function(y,th=c(1,.005,.6))
{return(c(th[1]*y[1], th[2]*y[1]*y[2],th[3]*y[2]))}

out=gillespie(N,10000)
op=par(mfrow=c(2,2))
plot(stepfun(out$t,out$x[,1]),pch="")
plot(stepfun(out$t,out$x[,2]),pch="")
plot(out$x,type='l')
par(op)
```


